---
title: "An Example of Using Function `estimateCounts`: Estimating the New Zealand Population"
author: "John Bryant"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example estimateCounts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

In this vignette, we use  function `estimateCounts` from package `demest` to estimate the population of New Zealand, by age, sex, and region, over the period 2008-2014, using simulated administrative data. The example is deliberately simplified--a real example would have more complicated data and models.

# Preparing and inspecting data

For the purposes of this example, we treat the official estimates of population, in array `nz.popn`, as the truth.  We use the various simulated datasets generated from `nz.popn` as data to try to recover the true population.  The datasets are as follows.

| Object in `demdata` | Description |
|--------------------:|:------------|
| nz.admin.nat | Population by age, sex, and year, 2008-2014 at the national level. Highly accurate. |
| nz.admin.health | Enrolments by age, sex, region, and year, 2008-2014.  Coverage varies by age.  |
| nz.admin.school | School enrolments by age, sex, region, and year, ages 5-14. Excellent proxy for population at these ages .|
| nz.admin.survey | Survey covering 5\% of the national population in 2010. |


The data are in package `demdata`.  The first step is to turn the data into counts arrays. 
```{r}
library(demest)
popn.true <- Counts(demdata::nz.popn.reg)
nat <- Counts(demdata::nz.admin.nat)
health <- Counts(demdata::nz.admin.health)
school <- Counts(demdata::nz.admin.school)
survey <- Counts(demdata::nz.admin.survey)
```

# Setting up the model

We begin by specifying a model for the underlying population.  The model is parsimonious, with the only interactions being between age and sex and between age and region, to help with convergence.

```{r}
popn.mod <- Model(y ~ Poisson(mean ~ age * sex + age * region + year),
                  region ~ Exch(error = Error(robust = TRUE)))
```

Next we specify data models and put them in a list.  The list does not need names, since `estimateCounts` will use the names of the responses to match data models to datasets.
```{r}
nat.mod <- Model(nat ~ PoissonBinomial(prob = 0.98))
health.mod <- Model(health ~ Poisson(mean ~ age))
school.mod <- Model(school ~ PoissonBinomial(prob = 0.98))
survey.mod <- Model(survey ~ Binomial(prob ~ 1))
observation <- list(nat.mod, health.mod, school.mod, survey.mod)
```

We generate some starting values for the population estimates,
```{r}
y <- Counts(array(as.integer(health * 1.2),
                  dim = dim(health),
                  dimnames = dimnames(health)))
```
put the datasest in a named list,
```{r}
datasets <- list(nat = nat, health = health, school = school, survey = survey)
```
and generate a filename
```{r}
filename <- tempfile()
```


# Running the model

We can now run the model. We will only do a few iterations, to see what happens.

```{r}
set.seed(1)  ## for replicability
estimateCounts(popn.mod,
               y = y ,
               observation = observation,
               datasets = datasets,
               filename = filename,
               nBurnin = 10,
               nSim = 10)
fetchSummary(filename)
```

Nothing has converged, which is to be expected, but the model did run.  After some experimentation, we change the jump lengths for the Metropolis-Hastings updates slightly,
to bring the acceptance rates closer to 40%.
```{r}
popn.mod <- Model(y ~ Poisson(mean ~ age * sex + age * region + year),
                  region ~ Exch(error = Error(robust = TRUE)),
		  jump = 0.08)
health.mod <- Model(health ~ Poisson(mean ~ age),
                    jump = 0.05)
survey.mod <- Model(survey ~ Binomial(prob ~ 1),
                    jump = 0.05)
observation <- list(nat.mod, health.mod, school.mod, survey.mod)
```
and rerun the model to convergence.  Warning: the next step can take a long time.
```{r, cache = TRUE}
set.seed(1)  ## for replicability
estimateCounts(popn.mod,
               y = y ,
               observation = observation,
               datasets = datasets,
               filename = filename,
               nBurnin = 5000,
               nSim = 5000,
               nThin = 20)
fetchSummary(filename)
```

# Examining the population estimates

To see what has been estimated, we use function `listContents`,
```{r}
listContents(filename)
```


To extract the population estimates, we use
```{r}
popn.est <- fetch(filename, where = "y")
```

We take a look at the results for the first year, 2008. The light blue bands show 95\% credible intervals and the dark blue bands show 50\% credible intervals.  Since we have the true values, we show them as well, in red.
```{r, fig.width = 7, fig.height = 7}
dplot(~ age | region,
      data = popn.est,
      subarray = year == "2008",
      prob = c(0.025, 0.25, 0.5, 0.75, 0.975),
      scales = list(y = "free"),
      main = "Population, by age and region, in 2008",
      overlay = list(values = subarray(popn.true, year == "2008"),
                     col = "red",
                     lwd = 2))
```

The same, for 2014,

```{r, fig.width = 7, fig.height = 7}
dplot(~ age | region,
      data = popn.est,
      subarray = year == "2014",
      prob = c(0.025, 0.25, 0.5, 0.75, 0.975),
      scales = list(y = "free"),
      main = "Population, by age and region, in 2014",
      overlay = list(values = subarray(popn.true, year == "2014"),
                     col = "red",
                     lwd = 2))
```

Changes in the Auckland population (which was relatively poorly estimated) over time, females only,
```{r, fig.width = 7, fig.height = 7}
dplot(~ year | age,
      data = popn.est,
      subarray = region == "Auckland" & sex == "Female",
      prob = c(0.025, 0.25, 0.5, 0.75, 0.975),
      main = "Auckland female population, by age and time",
      overlay = list(values = subarray(popn.true, region == "Auckland" & sex == "Female"),
                     col = "red",
                     lwd = 2))
```

Some numeric estimates
```{r}
nelson <- subarray(popn.est, region == "Nelson")
nelson <- collapseDimension(nelson, margin = "year")
nelson <- collapseIterations(nelson, prob = c(0.025, 0.5, 0.975))
round(nelson, -2)
```

Converted to a data.frame
```{r}
as.data.frame(nelson, direction = "long")
```

# Examining coverage levels

Among other things, the model generates 'coverage ratios', that is, the number of people in the dataset divided by estimates of the true population.  We take a look at coverage ratios for the health dataset in 2014.  Since we in fact have the true population, we show that too, in red.

```{r, fig.width = 7, fig.height = 7}
coverage.health.est <- fetch(filename, 
                         where = c("observation", "health", "likelihood", "mean"))
coverage.health.true <- health / popn.true
dplot(~ age | region,
      data = coverage.health.est,
      subarray = year == "2014",
      weights = popn.true,
      prob = c(0.025, 0.25, 0.5, 0.75, 0.975),
      main = "Coverage ratios, health data, by age and region, in 2014",
      overlay = list(values = subarray(coverage.health.true, year == "2014"),
                     col = "red",
                     lwd = 2))
```

